# Syntax and Functions {#Functions}

## Syntax

R is its own programming language, and some of the key syntax is not necessarily intuitive. We have covered some already, but here follows some important things to know:

Use `<-` to assign variables
e.g. `x <- 1`

A shortcut to producing is pressing the Alt + - keys together. 

Use `c()` to combine (i.e. to create) a vector. 
e.g. `c(1,2)`

Use `==` to evaluate something as being exactly equal to
```{r eval=FALSE}
1 = 1 #this will return an error
1 == 1 #this will return: [1] TRUE
```

Use `%in%` to specify that a character string or other data element is contained in some data structure.
```{r eval=FALSE}
1 %in% c(2, 2, 1) # this will return [1] TRUE as 1 is in the vector we've defined
```
Use `!=` to specify "is NOT equal to"
```{r eval=FALSE}
1 != 2 #this will return [1] TRUE as 1 is NOT equal to 2
```
Use :: to specify that a function comes from a specific package. This is necessary if you have two packages which include functions with the same name. 
```{r eval=FALSE}
# load the stringr package
library(stringr)
stringr::str_split("Let's split this string into 10 elements separated by spaces", " ") #this uses the str_split function from stringr
    [1] "Let's"     "split"     "this"      "string"    "into"      "10"        "elements"  "separated" "by"        "spaces"   
```

Use `%>%` from the Dplyr package - the pipe - to apply multiple functions to an object in turn. As words, the sequence would read "... and then". Use Ctrl + Shift + M as a keyboard shortcut to produce the pipe. 

```{r eval=FALSE}
library(dplyr)
y <- 25 #define y as the number 25
x <- y %>% #define a new element, x, as 25. Insert the pipe to continue the code chunk on our object, x.
  sqrt() %>%  # square root our element, x. *we do not need to specify that we're square rooting x as it's within a pipe
  print()   #print the object we've created then manipulated
[1] 5
```

Use `~` (tilde) to define the relationship between two sides of an equation. It separates the left and right side
```{r eval=FALSE}
two_by_two <- c(1:4) #this creates a vector of the numbers 1 to 4: 1,2,3,4
print(two_by_two)
[1] 1 2 3 4

Even_numbers <- case_when(two_by_two%%2 == 0 ~ "divisible by 2", # when the element in our two by two vector is divisible by 2, assign the text "divisible by 2"
          TRUE ~ "not divisible by 2")   #when our first condition is not met, assign "not divisible by 2
print(Even_numbers)
[1] "not divisible by 2" "divisible by 2"     "not divisible by 2" "divisible by 2"  
#this will create a vector of 4 character strings, which correspond to the two_by_two vector. Those in position 2 and 4 (the numbers 2 and 4) are divisible by 2, so returns "divisible by 2".
```


Use parentheses `()` to pass argumnts to a function

Use `$` to subset an object, similar to square brackets.

Use square brackets `[]` to grab a subset of data. It appends an index vector to a specified vector. Square brackets can usually be applied in place of `$`, and are more maleable to the specific task, though they might be slightly less intuitive. See the example below. For more detailed explanation of extracting/subsetting variables see the [data manipulation chapter](#extracting_variables)

```{r eval=FALSE}
data("iris") # load the iris data
unique(iris$Species) #show the unique entries in the Species column within the iris data frame
[1] setosa     versicolor virginica 

rep(unique(iris$Species), times = 2) # repeat x2 the unique entries from the Species column within the iris data frame
[1] setosa     versicolor virginica  setosa     versicolor virginica 

rep(unique(iris[[5]]), times = 2) #repeat x2 the unique entries from the 5th column within the iris data frame: the species column
[1] setosa     versicolor virginica  setosa     versicolor virginica 

```


Use `{}` to evaluate a series of items within the environment of the brackets. More on this when we discuss functions.


(Advanced) use `{{}}`, `:=` and `!!` when defining functions and needing to specify the nature of objects within them. More on this in [User-defined Functions chapter](#Overview) 


## BaseR Functions 

Functions are key to coding efficiently and effectively. BaseR, R's language, contains many functions. Packages also contain specific functions that improve on the inbuilt functionality of R. These functions speed up and optimise your code.

To see the function inputs, run ?function(). The help window will appear in the bottom right panel. You can also tab across from the open bracket once writing the function to see the arguments.

### Base R and Inbuilt Functions


`?()` get help on functions specified after the question mark. e.g. run `?rm()`

`rm()` removes a specified object from the global environment. This can also be applied to the whole environment with the brush icon.

`nrow()` how many rows are in a data frame

`nchar()` how many characters are in a string

`gsub(pattern, replace, x)` replaces matches in x with a string. Also see stringr and stringi packages for string manipulation.

`paste()` concatenate vectors

`paste0()`  concatenate strings

`is.na()` is missing

`!is.na()` is not missing

`is.null()` is null


### Programming functions

**If statements**

The most important thing to remember with if else statements is to make sure the else is on the same line as the close curly bracket of the if statement. Otherwise the else won't be tied to the preceding condition. 

```{r eval=FALSE}
if (condition){
  #Do something
} else {
  #Do something different
}

i <- 3
if (i > 3){
  print("yes")
} else {    #because the function is continuous, the else needs to be on the same line as the final curly bracket of the if function.
  print("no")
}

[1] "no"
```

Adding conditions to if statements using `else if `

```{r eval=FALSE}
i <- 2
if (i > 3){
  print("greater than 3")
} else if ( i == 3) {
  print("equal to 3")
} else {
  print("less than 3")
}

[1] "less than 3"
```


Loops are a good and easily comprehensible way to optimise your code, but take a bit of time to effectively apply to your work. More detail is contained in the [User-defined Functions](#Loops) chapter.

**For loop**
```{r eval=FALSE}
#Generic
for(variable in something) {
  do something
}

#Example application
for (i in 1:4){
  j <- i + 10
  print(j)
}

[1] 11
[1] 12
[1] 13
[1] 14

```

**While loop**

```{r eval=FALSE}
#Generic
while(condition){
  #do something
}

#Example application
i <- 1
while (i < 6) {
  print(i)
  i = i+1
}
[1] 1
[1] 2
[1] 3
[1] 4
[1] 5
```

## Quiz: R Syntax and Functions {#quiz}

```{r Syntax and functions quiz, echo=FALSE}
# Function to create and display a quiz on R syntax and functions
create_syntax_quiz <- function() {
  # A list containing all the quiz questions, options, and answers
  questions <- list(
    list(
      q = "How do you test if a value is 'exactly equal to' another in R?",
      options = c("A) <-", "B) =", "C) ==", "D) %in%"),
      answer = "C"
    ),
    list(
      q = "What is the purpose of the `%in%` operator?",
      options = c(
        "A) To perform integer division",
        "B) To check if an element is contained in a vector or list",
        "C) To multiply matrices",
        "D) To assign a value"
      ),
      answer = "B"
    ),
    list(
      q = "Which syntax is used to call a function that comes from a specific package?",
      options = c("A) package$function()", "B) function(package)", "C) package->function()", "D) package::function()"),
      answer = "D"
    ),
    list(
      q = "In the context of the `dplyr` package, what does the pipe operator `%>%` do?",
      options = c(
        "A) It subtracts one value from another",
        "B) It creates a plot",
        "C) It passes the object on its left to the function on its right",
        "D) It stops the execution of the code"
      ),
      answer = "C"
    ),
    list(
      q = "What is the primary use of the tilde `~` operator in R?",
      options = c(
        "A) To approximate a number",
        "B) To define the relationship between variables in a formula",
        "C) To negate a logical statement",
        "D) To specify a user's home directory"
      ),
      answer = "B"
    ),
    list(
      q = "Which function is used to remove a specified object from your R environment?",
      options = c("A) remove()", "B) delete()", "C) rm()", "D) clear()"),
      answer = "C"
    )
  )

  # Loop through each question and print it with its options
  for (i in seq_along(questions)) {
    cat("**", i, ". ", questions[[i]]$q, "**\n\n", sep = "")
    for (option in questions[[i]]$options) {
      cat(option, "\n")
    }
    cat("\n")
  }

  # Print a separator and the answers section header
  cat("---\n\n")
  cat("**Answers:**\n\n")

  # Loop through and print the correct answer for each question
  for (i in seq_along(questions)) {
    cat(i, ". ", questions[[i]]$answer, "\n", sep = "")
  }
}

# Run the function to generate the quiz
create_syntax_quiz()
```


